---
title: "Machine Learning"
author: Harald Puhr
date: "April 8, 2025"
---

[Data Source](https://github.com/ha-pu/data_files#repurchase-likelihood-dataset)

# Load packages and data

```{r}
library(caret)
library(caretEnsemble)
library(tidyverse)
```

Setting a seed is important for reproducibility. It ensures that the random
processes in your code yield the same results each time you run it. This is
particularly crucial in machine learning, where randomness can significantly
affect model training and evaluation.

```{r}
set.seed(123)
```

```{r}
df <- read_csv("https://raw.githubusercontent.com/ha-pu/data_files/refs/heads/main/4-repurchase_likelihood.csv")
```

# Preparing our data

```{r}
glimpse(df)
```

## Recode variables

```{r}
df <- df %>% 
  mutate(
    repurchase = factor(repurchase),
    customer_sex = factor(customer_sex),
    customer_tier = factor(customer_tier)
  )
```

## Drop columns which do not contain relevant information

```{r}
df <- select(df, -customer_id, -customer_name)
```

## Transform categorical predictors into dummy variables

```{r}
dummyModel <- dummyVars(repurchase ~ ., df, fullRank = TRUE)

df <- predict(dummyModel, newdata = df) %>%
  as_tibble() %>%
  mutate(repurchase = df$repurchase)
```

## Designate he column names of our outcome and predictor variables

```{r} 
label <- "repurchase"
predictors <- names(df)[!(names(df) %in% label)]
```

# Splitting the data into training and test sets

```{r}
train_idx <- sample(seq(nrow(df)), size = nrow(df) * 0.8)
train_df <- slice(df, train_idx)
test_df <- slice(df, -train_idx)
```

# Pre-processing the data

## Change data type of the label variable

```{r}
y_train <- train_df[,label][[1]]
y_test <- test_df[,label][[1]]

# Ensure levels of y_train are valid R variable names
levels(y_train) <- make.names(levels(y_train))
levels(y_test) <- make.names(levels(y_test))

head(y_train)
```

## Avoid multicollinearity

```{r}
findCorrelation(cor(train_df[,predictors]))
```

## Check for missing values

```{r}	
summary(train_df)
```

### Use median imputation

```{r}
preproc_par <- preProcess(train_df[,predictors], method = c("medianImpute"))
train_df[,predictors] <- predict(preproc_par, train_df[,predictors])
```

## Scale and center the data

```{r}
preproc_par <- preProcess(train_df[,predictors], method=c("center", "scale"))
train_df[,predictors] <- predict(preproc_par, train_df[,predictors])
```

```{r}
preprocPar <- preProcess(test_df[,predictors], method=c("center", "scale", "medianImpute"))
test_df[,predictors] <- predict(preprocPar, test_df[,predictors])
```

# Model training, validation, and testing

## Training

```{r, warning=FALSE, message=FALSE, include=FALSE}
quickTrain <- trainControl(method = "cv", number = 5)

model_rf <- train(x = train_df[,predictors], y = y_train, method = "rf", trControl = quickTrain)

model_glmnet <- train(x = train_df[,predictors], y = y_train, method = "glmnet", trControl = quickTrain)

model_nnet <- train(x = train_df[,predictors], y = y_train, method = "nnet", trControl = quickTrain)

model_xgbTree <- train(x = train_df[,predictors], y = y_train, method = "xgbTree", trControl = quickTrain)

model_xgbLinear <- train(x = train_df[,predictors], y = y_train, method = "xgbLinear", trControl = quickTrain)
```

## Validation

```{r}
comparison_alg <- tibble(
  Model = c("rf","glmnet","nnet","xgbTree","xgbLinear"),
  Accuracy_base = c(
    max(model_rf$results$Accuracy), max(model_glmnet$results$Accuracy),
    max(model_nnet$results$Accuracy), max(model_xgbTree$results$Accuracy),
    max(model_xgbLinear$results$Accuracy))
)
comparison_alg
```

### Additional validation methods

```{r, warning=FALSE, message=FALSE, include=FALSE}
# Cross validation training method
crossValidation <- trainControl(method = "CV")

# Boot strapping training method
bootstraping <- trainControl(method = "boot")

# Apply both methods to a random forest model
model_cv <- train(x =  train_df[,predictors], y = y_train, method = "rf", trControl = crossValidation)
model_boot <- train(x =  train_df[,predictors], y = y_train, method = "rf", trControl = bootstraping)
```

```{r}
comparison_val <- tibble(
  Model = c("cv","boot"),
  Accuracy_base = c(
    max(model_cv$results$Accuracy), max(model_boot$results$Accuracy)
  )
)
comparison_val
```

### Model tuning

```{r, warning=FALSE, message=FALSE, include=FALSE}
# Grid tuning
gridTune <- trainControl(method = "CV", search = "grid")

# Random search tuning method:
randomTune <- trainControl(method = "CV", search = "random")

# Adaptive search tuning method:
adaptiveTune <- trainControl(
  method = "adaptive_cv",
  adaptive = list(min = 2, alpha = 0.05, method = "gls", complete = TRUE),
  search = "random"
  )

# Apply all three methods to a glmnet model
model_grid <- train(
  x =  train_df[,predictors], y =  y_train, method = "glmnet",
  trControl = gridTune, tuneLength = 5
  )
model_random <- train(
  x =  train_df[,predictors], y =  y_train, method = "glmnet",
  trControl = randomTune, tuneLength = 25
  )
model_adaptive <- train(
  x =  train_df[,predictors], y =  y_train, method = "glmnet",
  trControl = adaptiveTune, tuneLength = 25
  )
```

```{r}
comparison_tune <- tibble(
  Model = c("grid", "random", "adaptive"),
  Accuracy_base = c(
    max(model_grid$results$Accuracy), max(model_random$results$Accuracy),
    max(model_adaptive$results$Accuracy)
  )
)
comparison_tune
```

## Testing

```{r}
comparison_all <- bind_rows(comparison_alg, comparison_val, comparison_tune)

det_OOS_accuracy <- function(model){
  confusionMatrix(predict(model, test_df), y_test)$overall[[1]]
}

all_model_list <- list(
  model_rf, model_glmnet, model_nnet, model_xgbTree,
  model_xgbLinear, model_cv, model_boot, model_grid,
  model_random, model_adaptive
)

comparison_all <- mutate(
  comparison_all,
  OOS_Acc = map_dbl(all_model_list, det_OOS_accuracy)
)

arrange(comparison_all, desc(OOS_Acc))
```

### Variable importance

```{r}
varImp(model_xgbTree)
```

```{r}
plot(varImp(model_xgbTree), main = "xgbTree")
```

```{r}
plot(varImp(model_nnet), main = "nnet")
```

```{r}
plot(varImp(model_glmnet), main = "glmnet")
```

# Ensemble learning

```{r, warning=FALSE, message=FALSE, include=FALSE}
algoList <- c("rf", "glmnet", "nnet", "xgbTree", "xgbLinear", "svmLinear2")

stack_index = createFolds(y_train, k = 5) # We want to fix our index for comparison and validity

listControl <- trainControl(
  method = "cv", number = 5, 
  classProbs = TRUE, # We need to predict probabilities not 0 or 1 (more info)
  savePredictions = TRUE, # necessary for stack model later
  index = stack_index
)

models <- caretList(
  x = train_df[,predictors], y = y_train, trControl = listControl,
  methodList = algoList
)
```

## Compare the models

```{r}
results <- resamples(models)
summary(results)
modelCor(resamples(models))
```

## Automatic model selection

```{r message=FALSE, warning=FALSE}
stackControl <- trainControl(
  method = "repeatedcv", number = 10, repeats = 3, classProbs = TRUE,
  index = stack_index
  )

stack_tree <- caretStack(models, method = "xgbTree", trControl = stackControl)

stack_gbm <- caretStack(models, method = "gbm", trControl = stackControl)

stack_glm <- caretStack(models, method = "glm", trControl = stackControl)
```

## Compare the models

```{r}
results <- tibble(
  ensemble_tree = predict(stack_tree, test_df, type = "prob"),
  ensemble_glm = predict(stack_glm, test_df, type = "prob"),
  ensemble_gbm = predict(stack_gbm, test_df, type = "prob")
)

results <- predict(models, test_df) %>%
  as_tibble() %>% 
  bind_cols(results)

det_OOS_accuracy_probs <- function(model){
  preds <- factor(model < 0.5, labels = c("X0", "X1"))
  confusionMatrix(preds, y_test)$overall[[1]]
}

OOS_ACC <- results %>%
  summarise(across(everything(), det_OOS_accuracy_probs))

OOS_ACC

OOS_ACC %>%
  pivot_longer(cols = everything(), names_to = "Model", values_to = "OOS_ACC") %>%
  ggplot() +
  geom_bar(aes(x = Model, y = OOS_ACC), stat = "identity") + 
  labs(
    x = NULL
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
