---
title: "Machine Learning Exercise"
author: Harald Puhr
date: "July 23, 2025"
---

[Data Source](https://github.com/ha-pu/data_files#subsidiary-income-dataset)

# Load packages and data

## Load packages

```{r}
library(caret)
library(tidyverse)
```

```{r}
set.seed(123)
```

## Load the data

```{r}
fdi_df <- read_csv(
    "https://raw.githubusercontent.com/ha-pu/data_files/refs/heads/main/5-subsidiary_income.csv",
    col_types = cols("i", "f", "i", "i", "n", "n", "n", "n", "n", "n", "n", "f", "f", "f")
) %>%
    drop_na()
```

## Take a look at the data

```{r}
glimpse(fdi_df)
```

# Prepare the data


## Lagging predictors

We need need to create predictions on the data from the previous years, so we
have to lag the data.

```{r}
lagged_df <- fdi_df %>%
    arrange(Year) %>%
    mutate(
        across(.cols = !contains("_L"), .fns = list(L1 = ~ lag(., n = 1))),
        across(.cols = !contains("_L"), .fns = list(L2 = ~ lag(., n = 2))),
        .by = c(DomesticId, ForeignId)
    ) %>%
    select(
        !contains("Year_L") &
            !contains("CountryCode_L") &
            !contains("DomesticId_L") &
            !contains("ForeignId_L")
    ) %>%
    replace_na(list(Income = 0))
```

## Remove features

Now we need to remove the features we cannot know in advance.

```{r}
data_df <- select(
    lagged_df,
    -c(
        "NumFeature1", "NumFeature2", "NumFeature3", "NumFeature4",
        "NumFeature5", "NumFeature6", "FactorFeature1", "FactorFeature2",
        "FactorFeature3"
    )
)
```

And finally, we also remove all observations for which we do not have any
information and drop the Id columns.

```{r}
data_df$na_check <- rowSums(is.na(data_df[6:ncol(data_df)]))

data_df <- data_df %>%
    filter(na_check != 30) %>%
    select(-c(na_check, Year, CountryCode, DomesticId, ForeignId)) %>%
    as.data.frame()
```

## Take a look at the data

```{r}
glimpse(data_df)
```

# Modeling setup

## Transform categorical predictors into dummy variables

```{r}
dummyModel <- dummyVars(Income ~ ., data_df, fullRank = FALSE)

data_df <- data_df %>%
    select(Income) %>%
    bind_cols(data.frame(predict(dummyModel, newdata = data_df)))
```

## Define dependent and independent variables

Let's define which column is our DV and which are the IVs

```{r}
label <- "Income"
predictors <- setdiff(names(data_df), label)
```

## Train-test split

Now we have to create a train test split

```{r}
train_index <- createDataPartition(data_df$Income, p = 0.8, list = FALSE)

train_df <- data_df[train_index, ]

test_df <- data_df[-train_index, ]
```

## Preprocessing

Now we can decide on how to preprocess our data. Since we have a lot of dummies,
it is likely helpful to drop low variance columns. Also we have many missing 
values so we should use some imputation methods. 

```{r}
pre_process <- (method <- c("medianImpute", "nzv"))
```

## Define training parameters

For the next step we have to decide on a train control method.
To keep things consistent we want to manually define our folds first.

```{r}
testFolds <- createFolds(train_df[, label], k = 2)
```

Set training control to cross validation and adaptive tuning

```{r}
train_control <- trainControl(
    method = "cv",
    index = testFolds,
    verboseIter = TRUE
)
```

# Model training

First let's train a gradient boosted tree model

```{r}
model_xgbTree <- train(
    x = train_df[, predictors],
    y = train_df[, label],
    method = "xgbTree",
    tuneLength = 2,
    trControl = train_control,
    preProcess = pre_process
)
```

# Testing

As the last step we need to see how well our models are doing.
First we need to use the model to create predictions for our test set.

```{r}
test_df$pred_xgbTree <- predict(model_xgbTree, newdata = test_df)
```

Lets plot and see how well we do:

```{r}
test_df %>%
    ggplot() +
    geom_point(aes(x = Income, y = pred_xgbTree)) +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    geom_abline(slope = cor(test_df$Income, test_df$pred_xgbTree), intercept = 0, color = "blue") +
    labs(
        x = "Actual",
        y = "Predicted",
        title = paste("Correlation: ", round(cor(test_df$Income, test_df$pred_xgbTree), 3), sep = "")
    ) +
    theme_bw()
```

# Exercise

Use the code from `4-machine_learning.rmd` to train other models on the same
data. Compare how your models are doing vis-à-vis the `xgbTree` model.

You can find a summary of all models included in the `caret` package
[here](https://topepo.github.io/caret/available-models.html).
